A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-28 LyapDraft.tex
small sign error
minor change
Update LyapDraft.tex

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratics costs, Polyak was able to apply an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's Iterative Method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex functions. Indeed, Lessard~\emph{et al.} derived a simple one dimensional counterexample where the standard Heavy Ball method does not converge~\cite{Lessard14}.

In order to make momentum methods rigorous, a different approach was required.  In celebrated work, Nesterov devised a general scheme to accelerate convex optimization methods, which achieve optimal running times for a myriad of oracle models in convex programming~\cite{Nesterov04}. However, to achieve such rigorous general applicability, Nesterov's proof techniques abandoned the physical intuition of Polyak~\cite{Nesterov04}.  In lieu of differential equations and potential functions, Nesterov devised the method of \emph{estimate sequences} to verify the correctness of these momentum-based methods.  Researchers have struggled with understanding the intuition and underpinnings of the estimate sequence methodology since Nesterov's initial papers.  Nesterov himself often refers to the associated proof techniques as an ``algebraic trick.''

To overcome the lack of intuition for the estimate sequences technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  Orthogonally, another set of authors have taken a different approach by analyzing the continuous-time limit of accelerated methods and showing that the stability of the resulting ODEs can be verified by analyzing a simple Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  Unfortunately, none of this previous work provides a clear path of obtaining a discrete time optimization algorithm from a continuous time ODE.  There are a vast number of ways to discretize ODEs, but not all of them give rise to convergent methods.  Indeed, for unconstrained optimization on Euclidean space, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent in the setting where the objective is strongly convex have the same continuous time limit.

@@

\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k),  \label{Eq:ZSeqAcc3}\\%\label{Eq:ZSeqAcc2}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(y_{k+1})  \label{Eq:XSeqAcc3}\\ %\label{Eq:XSeqAcc}\\
\quad M \|L^{-1} \nabla f(y_{k})\|_\ast^{2} &\leq L^{-1}\langle\nabla f(y_{k}), x_{k} - y_{k}\rangle \label{Eq:YSeqAcc3}
\end{align}
\end{subequations}
The advantage of this version of the accelerated mirror descent algorithm is that it can be extended to accelerate higher-order gradient methods (the details of which we give in Appendix~\ref{App:ProofAccGrad}). The second version, originally introduced by Nesterov~\cite{Nesterov05} entails computing the mirror descent update~\eqref{Eq:ZSeqAcc2} from the gradient at $x$:
\begin{equation*}
z_k = \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla  f(x_i), z\rangle + D_h(z, z_0)\right\}
\end{equation*}
which results in an algorithm with the following optimality conditions, 
\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k), \label{Eq:ZSeqAcc4}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(x_{k+1})\label{Eq:XSeqAcc4}\\
\quad f(y_k) - f(x_k)  &\leq  -\frac{1}{2L} \| \nabla f(x_{k})\|_\ast^{2} \label{Eq:YSeqAcc4}
\end{align}
\end{subequation}
As mentioned above, to obtain a convergence guarantee we have replaced the current state $x_k$ by another sequence $y_k$ which simply needs to satisfy~\eqref{Eq:YSeqAcc3}. Furthermore, we have replaced our sequence $A_{k+1}$ by $\tilde A_{k+1}$ which depends on the Lipschitz parameter. We make the same adjustments to the Lyapunov function~\eqref{Eq:DiscLyapFunc1} which we sum up in the following theorem:
\begin{theorem} \label{Thm:AccGrad}
Using the following Lyapunov function
\begin{equation}\label{Eq:LyapAccGrad}
E_k = \tilde A_k(  f(y_k) -  f(x)) + D_h(x, z_k),
\end{equation}
we can show
  % which has the effect of scaling the gradient by $\epsilon$ as well. 
\[E_{k+1} - E_k \leq \varepsilon_k\]
where the error scales in the following way 
\[\varepsilon_k =  \left(\frac{1}{2}(\tilde A_{k+1} -\tilde A_{k})^{2}- \tilde A_{k+1}M\right)\|\nabla f(y_{k+1})\|^2 \]
for the accelerated gradient method~\eqref{Eq:AcceleratedGrad2}.
\end{theorem}
Choosing $x = x^\ast$ and $A_k$ such that $\tilde A_{k+1}^{-1}(\tilde A_{k+1} -\tilde A_{k})^{2} \leq 2M$ so that the error is non-positive, we obtain the following convergence rate,
\begin{equation}
 f(y_k) -  f(x^\ast) \leq \frac{\tilde A_0(f(x_0) - f(x^\ast)) + D_h(x^\ast, x_0)}{\tilde A_k}.
\end{equation}
Note, this condition requires that $A_k$ can grow at most quadratically which gives an $O(1/k^2)$ rate of convergence.
We give a proof for both of these methods in Appendix~\ref{App:ProofAccGrad}

@@

\subsection{Accelerated Gradient Descent: Version 2}
We now analyze the second version of the accelerated gradient descent algorithm using the same energy functional~\eqref{Eq:LyapAccGrad}:
\begin{align*}
E_{k+1} - E_k &= \tilde A_{k+1}(f(y_{k+1}) - f(x^\ast)) - \tilde A_{k}(f(y_{k}) - f(x^\ast)) + D_h(x^\ast, z_{k+1}) - D_h(x^\ast, z_{k}) \\
& = \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad - \langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1}\rangle + D_h(z_{k+1}, z_k)\\
& \overset{\eqref{Eq:XSeqAcc4}}{\leq} \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k+1}\rangle - \frac{1}{2}\|z_{k+1} - z_k\|^2\\
& = \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}),  z_{k} - z_{k+1}\rangle- \frac{1}{2}\|z_{k+1} - z_k\|^2\\
& \leq \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \overset{\eqref{Eq:ZSeqAcc4}}{=}\tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k+1} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - y_k\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& =\tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \overset{\eqref{Eq:YSeqAcc4}}{\leq}\frac{\tilde A_{k+1}}{2L}\|\nabla f(x_{k+1})\|^2 + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \leq \frac{1}{2L^2}\left((A_{k+1} -  A_k)^2 - A_{k+1} \right) \|\nabla f(x_{k+1})\|^2
\end{align*}
Therefore we have shown 
\begin{align*}
E_{k+1} - E_k \leq \varepsilon_k 
\end{align*}
where 
\begin{align*}
\varepsilon_k = \frac{1}{2L^2}\left[(A_{k+1} -  A_k)^2 - A_{k+1} \right] \|\nabla f(x_{k+1})\|^2
\end{align*}
To ensure our error $\varepsilon_k$ is non-positive, $A_k$ can be at most a polynomial of degree 2. %Note, this matches the bound for our theorem with $M = \frac{1}{2}$ (or, $N = 1$)

>>>>

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratic costs, Polyak was able to provide an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's Iterative Method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex functions. Indeed, Lessard~\emph{et al.} derived a simple one-dimensional counterexample where the standard Heavy Ball method does not converge~\cite{Lessard14}.

In order to make momentum methods rigorous, a different approach was required.  In celebrated work, Nesterov devised a general scheme to accelerate convex optimization methods, achieving optimal running times under oracle models in convex programming~\cite{Nesterov04}. However, to achieve such general applicability, Nesterov's proof techniques abandoned the physical intuition of Polyak~\cite{Nesterov04}; in lieu of differential equations and potential functions, Nesterov devised the method of \emph{estimate sequences} to verify the correctness of these momentum-based methods.  Researchers have struggled with understanding the intuition and underpinnings of the estimate sequence methodology since Nesterov's initial papers.  The associated proof techniques are often viewed as an ``algebraic trick.''

To overcome the lack of intuition for the estimate sequences technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  One promising general approach to the analysis of acceleration has been to analyze the continuous-time limit of accelerated methods and to show that the stability of the resulting ODEs can be verified via a Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  Unfortunately, none of this previous work provides a clear path of obtaining a discrete time optimization algorithm from a continuous time ODE.  There are a vast number of ways to discretize ODEs, but not all of them give rise to convergent methods.  Indeed, for unconstrained optimization on Euclidean space, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent in the setting where the objective is strongly convex have the same continuous time limit.

@@

\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k),  \label{Eq:ZSeqAcc3}\\%\label{Eq:ZSeqAcc2}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(y_{k+1})  \label{Eq:XSeqAcc3}\\ %\label{Eq:XSeqAcc}\\
\quad M \|L^{-1} \nabla f(y_{k})\|_\ast^{2} &\leq L^{-1}\langle\nabla f(y_{k}), x_{k} - y_{k}\rangle \label{Eq:YSeqAcc3}
\end{align}
\end{subequations}
The advantage of this version of the accelerated mirror descent algorithm is that it can be extended to accelerate higher-order gradient methods (the details of which we give in Appendix~\ref{App:ProofAccGrad}). The second version, originally introduced by Nesterov~\cite{Nesterov05} entails computing the mirror descent update~\eqref{Eq:ZSeqAcc2} from the gradient at $x$:
\begin{equation*}
z_k = \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla  f(x_i), z\rangle + D_h(z, z_0)\right\}
\end{equation*}
which results in an algorithm with the following optimality conditions, 
\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k), \label{Eq:ZSeqAcc4}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(x_{k+1})\label{Eq:XSeqAcc4}\\
\quad f(y_k) - f(x_k)  &\leq  -\frac{1}{2L} \| \nabla f(x_{k})\|_\ast^{2} \label{Eq:YSeqAcc4}
\end{align}
\end{subequation}
As mentioned above, to obtain a convergence guarantee we have replaced the current state $x_k$ by another sequence $y_k$ which simply needs to satisfy~\eqref{Eq:YSeqAcc3}. Furthermore, we have replaced our sequence $A_{k+1}$ by $\tilde A_{k+1}$ which depends on the Lipschitz parameter. We make the same adjustments to the Lyapunov function~\eqref{Eq:DiscLyapFunc1} which we sum up in the following theorem:
\begin{theorem} \label{Thm:AccGrad}
Using the following Lyapunov function
\begin{equation}\label{Eq:LyapAccGrad}
E_k = \tilde A_k(  f(y_k) -  f(x)) + D_h(x, z_k),
\end{equation}
we can show
  % which has the effect of scaling the gradient by $\epsilon$ as well. 
\[E_{k+1} - E_k \leq \varepsilon_k\]
where the error scales in the following way 
\[\varepsilon_k =  \left(\frac{1}{2}(\tilde A_{k+1} -\tilde A_{k})^{2}- \tilde A_{k+1}M\right)\|\nabla f(y_{k+1})\|^2 \]
for the accelerated gradient method~\eqref{Eq:AcceleratedGrad2}.
\end{theorem}
Choosing $x = x^\ast$ and $A_k$ such that $\tilde A_{k+1}^{-1}(\tilde A_{k+1} -\tilde A_{k})^{2} \leq 2M$ so that the error is non-positive, we obtain the following convergence rate,
\begin{equation}
 f(y_k) -  f(x^\ast) \leq \frac{\tilde A_0(f(x_0) - f(x^\ast)) + D_h(x^\ast, x_0)}{\tilde A_k}.
\end{equation}
Note, this condition requires that $A_k$ can grow at most quadratically which gives an $O(1/k^2)$ rate of convergence.
We give a proof of both of these methods in Appendix~\ref{App:ProofAccGrad}.

@@

\subsection{Accelerated Gradient Descent: Version 2}
We now analyze the second version of the accelerated gradient descent algorithm using the same energy functional~\eqref{Eq:LyapAccGrad}:
\begin{align*}
E_{k+1} - E_k &= \tilde A_{k+1}(f(y_{k+1}) - f(x^\ast)) - \tilde A_{k}(f(y_{k}) - f(x^\ast)) + D_h(x^\ast, z_{k+1}) - D_h(x^\ast, z_{k}) \\
& = \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad - \langle \nabla h(z_{k+1}) - \nabla h(z_k), x^\ast - z_{k+1}\rangle - D_h(z_{k+1}, z_k)\\
& \overset{\eqref{Eq:XSeqAcc4}}{\leq} \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k+1}\rangle - \frac{1}{2}\|z_{k+1} - z_k\|^2\\
& = \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}),  z_{k} - z_{k+1}\rangle- \frac{1}{2}\|z_{k+1} - z_k\|^2\\
& \leq \tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - z_{k}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \overset{\eqref{Eq:ZSeqAcc4}}{=}\tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k+1} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - y_k\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& =\tilde A_{k+1}(f(y_{k+1}) - f(x_{k+1})) + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \overset{\eqref{Eq:YSeqAcc4}}{\leq}\frac{\tilde A_{k+1}}{2L}\|\nabla f(x_{k+1})\|^2 + \tilde A_{k}(f(x_{k+1}) - f(y_{k})) + (\tilde A_{k+1} - \tilde A_k)(f(x_{k+1}) - f(x^\ast)) \\
&\quad  + \tilde A_{k} \langle \nabla f(x_{k+1}),  y_k - x_{k+1}\rangle + (\tilde A_{k+1} - \tilde A_k)\langle \nabla f(x_{k+1}), x^\ast - x_{k+1}\rangle + \frac{(\tilde A_{k+1} - \tilde A_k)^2}{2}\|f(x_{k+1})\|^2 \\
& \leq \frac{1}{2L^2}\left((A_{k+1} -  A_k)^2 - A_{k+1} \right) \|\nabla f(x_{k+1})\|^2
\end{align*}
Therefore we have shown 
\begin{align*}
E_{k+1} - E_k \leq \varepsilon_k 
\end{align*}
where 
\begin{align*}
\varepsilon_k = \frac{1}{2L^2}\left[(A_{k+1} -  A_k)^2 - A_{k+1} \right] \|\nabla f(x_{k+1})\|^2
\end{align*}
To ensure our error $\varepsilon_k$ is non-positive, $A_k$ can be at most a polynomial of degree 2. %Note, this matches the bound for our theorem with $M = \frac{1}{2}$ (or, $N = 1$)
